Programmatic Generation of R Markdown Documents with \pkg{listdown}

### 2020-04-28 inst/r-journal/listdown-article.rmd
Major bigtime revision.

Since computational components are, by definition, computationally derived objects and R Markdown is a well-defined standard it is possible to programmatically create R Markdown documents with computational components. This is the focus of this paper. However, before proceeding down this avenue, we would like to highlight two fundamental limitations to automated R Markdown document construction. First and foremost, without narrative components a document has no context. Quantitative analyses require research questions, hypotheses, reviews, interpretations, and conclusions. Computational components are necessary but not sufficient for constructing an analysis. Second, it is not possible to construct computational components for an arbitary set of distinct analyses. An analysis itself has a context and it is built with a set of assumptions and goals. Computational components are constructed for a narrow class of problems.

However, this is not to say the programmatic creation of R Markdown documents is unwarranted or not useful. In fact it has at least two appealing characteristics. The first is convenience. <FILL THIS IN - TALK ABOUT REUSE TO MOTIVATE LARGER PIPELINES>. The second is that for fixed analysis pipelines processing uniformly-formatted but domain-distinct data, automated document generation enforces uniformity across those domains being serviced. <FILL THIS IN USE MULTIPLE-HISTOLOGY THERAPY AS AN EXAMPLE>.

The \pkg{listdown} package provides functions to programmatically create R Markdown files from 
named lists. It is intended for data analysis pipelines where the presentation of the results
is separated from their creation. For this use case, a data processing (or analysis) is performed
and the results are provided in a single named list, organized heirarchically. With the list and a \pkg{listdown} object a workflowr [@blischak2019], pdf, word, or html page. List element names denote sections, subsections,
subsubsection, etc. and the list elements contain the data structure to be presented including
graphs and tables. The goal of the package is not to provide a finished, readable document. It is to provide a document with all tables and visualization that will appear (_computational_ components). This serves as a starting point from which a user  can organize outputs, describe a study, discuss results, and provide conclusions (_narrative_ components).

@@

The package itself is relatively simple with 6 distinct methods that can be easily incorporated into existing analysis pipelines for automatically creating documents that can be used for data exploration and reviewing analysis results as well as a starting point for a more formal write up. These methods include:

\begin{itemize}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf listdown() }{- create a listdown object to create an R Markdown document}
\end{itemize}

@@

Suppose we have just completed and analysis and have collected all of the results into a list where the list elements are roughly in the order we would like to present them in a document. It may be noted that this is not always how computational components derived from data analyses are collated. Often individual components are stored in multiple locations on a single machine or across machines. However, it is important to realize that even for analyses on large-scale data, the digital artifact that will be presented will be relatively small. Centralizing them makes it easier to access them, since they don't need to be found in mulitple locations. Also, storing them as a list provides a hierarchical structure that translates directly to a document as we will see below.

@@

Creating a document from the `computational_components` will require two steps. First, we will create a `listdown` object that specifies how the `computational_components` object will be loaded into the document,  which libraries and code needs to be included, the options for the R chunks, and how the list elements will be presented in the output R markdown document.

@@

ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"), # The expression to load pres_list in the .Rmd.
               package = "ggplot2")                     # The packges needed to render plots.
```

The `ld` object, along with the computational components in the `comp-comp.rds` file are sufficient to to create the sections, subsections, and R chunks of a document. The only other thing requires to create the document is the header. The listdown package currently supports regular R Markdown, workflowr, and ioslides headers. A complete document can then be written to the console using the code shown below. It could easily be written to a file for rendering using the `writeLines()` function, for example.

```{r}
doc <- c(
  ld_rmarkdown_header("Anscombe's Quartet",
                     author = "Francis Anscombe",
                     date = "1973"),
  ld_make_chunks(ld))

@@

The first example is simple in part because the ggplot objects both contain the data we want to display and, at the same time, provide the mechanism for presenting them - rendering them in a graph. However, this is not always the case. The objects being stored in the list of computational components may not translate directly to the presentation in a document. In these cases a function is needed that takes the list component and returns an object to be displayed. For example, suppose that, along with showing graphs from the Anscombe Quartet, we would like to include the data themselves. We could add the data to the `computational_components` list and then create the document with:

@@

In \pkg{listdown} a function or method that implments the presentation of a computational component is referred to as a _decorator_ since if follows the classic decorator pattern described in @gamma1995. A decorator takes the element that will be presented as an argument and returns an object for presentation in the output directory. A decorator is specified using the `decorator` parameter of the `listdown()` function using a named list where the name corresponds to the type and the element correspond to the function or method that will decorate an object of that type. For example, the `anscombe` data set can be decorated with the `DT::datatable()` function [@xie2020] as:

@@

List names in the `decorator` argument provide a key to which a function or method is mapped. The underlying decorator resolution is implemented for a given computational component by going through decorator names sequentially to see if the component inherits from the name using the `inherits()` function. The function or method is selected from the corresponding name which the element first inherits from. This means that when customizing the presentation of objects that inherit from a common class, the more abstract classes should appear at the end of the list. This will ensure that specialized classes will be encountered first in the resolution process.

@@

# Advanced usage

Along with the basic usage, the \pkg{listdown} package supports methods to further customize the presentation. Two specific areas of customization are addressed. The first is where custom code is required either to modify decorators or even define the decorators. In this case, we would to add a small portion of code, immediately after loading required libraries. In the second case, rather than changing the R code inside of a chunk, we would like to change the options for an individual chunk. Both of these features are supported in the package and are presented below.

@@

```{r}
saveRDS(comp_comp2, "comp-comp2.rds")
ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
               package = c("ggplot2", "DT", "purrr"),
               decorator = list(ggplot = identity,
                                data.frame = datatable),
               init_expr = {
                 datatable <- partial(datatable,
                                      options = list(dom = 't'))
                 },
               echo = FALSE)

@@

```{r}
comp_comp2$Iris <- ld_chunk_opts(comp_comp2$Iris, echo = TRUE)
saveRDS(comp_comp2, "comp-comp2.rds")
cat(paste(ld_make_chunks(ld)[1:19], collapse = "\n"))
```
# Conclusion

>>>>

Since computational components are, by definition, computationally derived objects and R Markdown is a well-defined standard it is possible to programmatically create R Markdown documents with computational components. This is the focus of this paper. However, before proceeding down this avenue, we would like to highlight two fundamental limitations to automated R Markdown document construction. First and foremost, without narrative components a document has no context. Quantitative analyses require research questions, hypotheses, reviews, interpretations, and conclusions. Computational components are necessary but not sufficient for constructing an analysis. Second, it is not possible to construct computational components for an arbitary analyses. Analyses themselves have context and are built with a set of assumptions and goals. Computational components are constructed for a narrow class of problems.

However, this is not to say the programmatic creation of R Markdown documents is unwarranted or not useful. In fact, it has at least two appealing characteristics. The first is that it allows us to distinguish the presentation of analytical results from other steps in a data science or data processing pipeline. These other steps including cleaning and analysis often require their own environment and configuration with requirements very different than the computational needs of creating a presentation. By separating these components each can be developed independently. At the same time, by specifying that the output of those objects are organized hierarchically, we are essentially creating a contract by which processed data can be passed to the presentation as well as providing a way to organize the results. The second reason for programmatic creation of R Markdown documents is convenience. In collaborative environments, especially in the early stages, large numbers of graphs and tables are generated and discussed. By collecting these artifacts hierarchicaly as a list, they can be quickly iterated upon and restructured to more clearly present the data without needing to spend time on the creation of the presentation document.

The \pkg{listdown} package provides functions to programmatically create R Markdown files from 
named lists. It is intended for data analysis pipelines where the presentation of the results
is separated from their creation. For this use case, a data processing (or analysis) is performed
and the results are provided in a single named list, organized heirarchically. With the list and a \pkg{listdown} object a \pkg{workflowr} [@blischak2019], pdf, word, or html page. List element names denote sections, subsections,
subsubsection, etc. and the list elements contain the data structure to be presented including
graphs and tables. The goal of the package is not to provide a finished, readable document. It is to provide a document with all tables and visualization that will appear (_computational_ components). This serves as a starting point from which a user  can organize outputs, describe a study, discuss results, and provide conclusions (_narrative_ components).

@@

One area where we have found \pkg{listdown} is particuarly useful is in the reporting and research of clinical trial data. These collaborations tend to be between (bio)statisticians and clinicians either analyzing past trial data to formulate a new trial or in trial monitoring where trial telemetry (enrollment, responses, etc.) is reported and initial analyses are conveyed to a clinician. The associated presentations require very little context - clinicians often have a better understanding of the data collected than the statistician - often eliminating the need for narrative components. At the same time, a large number of hierarchical, heterogenous artifacts (tables and multiple types of plots) need to be conveyed thereby making the manual creation of R Markdown documents inconvenient.

Furthermore, data processing can be fixed across trials. This is especially true in the initial stages, which focus on patient demographics and enrollment. This has made it convenient for our group to quickly generate standardize and complete reports for multiple trials concurrently. To date, we have used listdown to report on five clinical trials, with another two currently in process. Results are disseminated using the \pkg{workflowr} package, usually with nine tabs conveying aspects of the data from collection through several different analyses, and each tab containing approxiately five to thirty tables, plots, or other artifacts including \pkg{trelliscopejs} [@trelliscopejs] environments which may hold hundreds of graphs. By generating many presentation artifacts we are able to address data-drive questions and issues during collaborative sessions and by carefully structuring these elements we allowing all members to participate in the process.

The \pkg{listdown} package itself is relatively simple with 6 distinct methods that can be easily incorporated into existing analysis pipelines for automatically creating documents that can be used for data exploration and reviewing analysis results as well as a starting point for a more formal write up. These methods include:

\begin{itemize}
\item{\bf as\_ld\_yml() }{- turn a computational component list into YAML with class information}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf listdown() }{- create a listdown object to create an R Markdown document}
\end{itemize}

@@

Suppose we have just completed and analysis and have collected all of the results into a list where the list elements are roughly in the order we would like to present them in a document. It may be noted that this is not always how computational components derived from data analyses are collated. Often individual components are stored in multiple locations on a single machine or across machines. However, it is important to realize that even for analyses on large-scale data, the digital artifacts to be presented are relatively small. Centralizing them makes it easier to access them, since they don't need to be found in mulitple locations. Also, storing them as a list provides a hierarchical structure that translates directly to a document as we will see below.

@@

Creating a document from the `computational_components` will require two steps. First, we will create a `listdown` object that specifies how the `computational_components` object will be loaded into the document,  which libraries and code needs to be included, and how the list elements will be presented in the output R markdown document.

@@

ld <- listdown(load_cc_expr = readRDS("comp-comp.rds"),
               package = "ggplot2")
```

The `ld` object, along with the computational components in the `comp-comp.rds` file are sufficient to to create the sections, subsections, and R chunks of a document. The only other thing requires to create the document is the header. The listdown package currently supports regular R Markdown and \pkg{workflowr} as \code{yml} objects from the \pkg{yaml} package [@yaml]. These objects are stored as named lists in R and are easily modified to accomodate document parameters. A complete document can then be written to the console using the code shown below. It could easily be written to a file for rendering using the `writeLines()` as in the following example.

```{r}
doc <- c(
  as.character(
    ld_rmarkdown_header("Anscombe's Quartet",
                        author = "Francis Anscombe",
                        date = "1973")),
  ld_make_chunks(ld))

@@

The first example is simple in part because the ggplot objects both contain the data we want to display and, at the same time, provide the mechanism for presenting them - rendering them in a graph. However, this is not always the case. The objects being stored in the list of computational components may not translate directly to the presentation in a document. In these cases, a function is needed that takes the list component and returns an object to be displayed. For example, suppose that, along with showing graphs from the Anscombe Quartet, we would like to include the data themselves. We could add the data to the `computational_components` list and then create the document with:

@@

In \pkg{listdown}, a function or method that implements the presentation of a computational component is referred to as a _decorator_ since if follows the classic decorator pattern described in @gamma1995. A decorator takes the element that will be presented as an argument and returns an object for presentation in the output directory. A decorator is specified using the `decorator` parameter of the `listdown()` function using a named list where the name corresponds to the type and the element correspond to the function or method that will decorate an object of that type. For example, the `anscombe` data set can be decorated with the `DT::datatable()` function [@xie2020] as:

@@

List names in the `decorator` argument provide a key to which a function or method is mapped. The underlying decorator resolution is implemented for a given computational component by going through decorator names sequentially to see if the component inherits from the name using the `inherits()` function. The function or method is selected from the corresponding name which the element first inherits from. This means that when customizing the presentation of objects that inherit from a common class, the more abstract classes should appear at the end of the list. This will ensure that specialized classes will be encountered first in the resolution process. It should be noted that an object's type is first checked against the decorator name list and then checked to see if it is a list. This allows a user to both decorate a list and retain `"list"` in its class attributes.

@@

# R code chunk customization

Along with the basic usage, the \pkg{listdown} package supports two other capabilities to further customize the presentation. The first is where custom code is required either to modify decorators or even define a decorators. In this case, we would to add a small portion of code, immediately after loading required libraries. In the second case, rather than changing the R code inside of a chunk, we would like to change the options for an individual chunk. Both of these features are supported in the package and are presented below.

@@

The `listdown()` function facilitates the insertion of initialization code through the `init_expr` argument. Code is inserted immediately after the libraries are loaded in the R Markdown document. In general, it is suggested that the number of initial expressions be kept small so that the R Markdown document is easy to read. If a large number of functions are required by the target R Markdown document then they can be put into a file and sourced using the initial expression. As an example, suppose we are creating an html document and presenting data using the `datatable()` function. However, we do not want to include the search capabilities provided. This can be easily accomplished by creating a new function, `datatable_no_search()`, created using the `partial()` function \cite{purrr} to partially apply `list(dom = 't')` to the `options` argument of `datatable`.

```{r}
saveRDS(comp_comp2, "comp-comp2.rds")
ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
               package = c("ggplot2", "DT", "purrr"),
               decorator = list(ggplot = identity,
                                data.frame = datatable_no_search),
               init_expr = {
                 datatable_no_search <- partial(datatable,
                                                options = list(dom = 't'))
                 },
               echo = FALSE)

@@

```{r}
comp_comp2$Iris <- ld_chunk_opts(comp_comp2$Iris, echo = TRUE)
saveRDS(comp_comp2, "comp-comp2.rds")
cat(paste(ld_make_chunks(ld)[1:19], collapse = "\n"))
```

# A simple example: Reporting on the `gtsummary::trial` dataset

\label{sect:simple-example}

```{r, eval = FALSE}
library(gtsummary)
library(dplyr)
library(glue)
library(survival)
library(survminer)
library(rmarkdown)

# Function to create the computational components.
make_surv_cc <- function(trial, treat, surv_cond_chars) {
  # Create Table 1 by treatment
  table_1 <- trial %>%
    tbl_summary(by = all_of(treat)) %>%
    gtsummary::as_flextable()

  # Create survival plots base on survival characteristics
  scs <- lapply(c("1", surv_cond_chars),
                function(sc) {
                  glue("Surv(ttdeath, death) ~ {treat} + {sc}") %>%
                    as.formula() %>%
                    surv_fit(trial) %>%
                    ggsurvplot()
                })
  names(scs) <- c("Overall", tools::toTitleCase(surv_cond_chars))
  list(`Table 1` = table_1, `Survival Plots` = scs)
}

# Create the computational components and write to surv-cc.rds.
make_surv_cc(trial, treat = "trt",
             surv_cond_chars = c("stage", "grade")) %>%
  saveRDS("surv-cc.rds")

# Create the listdown object with appropriate dependencies
# and presentation parameters.
ld_surv <- listdown(load_cc_expr = readRDS("surv-cc.rds"),
                    package = c("gtsummary", "survminer"),
                    decorator = list(ggsurvplot = identity,
                                     tbl_summary = identity),
                    echo = FALSE,
                    message = FALSE,
                    warning = FALSE,
                    fig.width = 7,
                    fig.height = 4.5)

# Write an RMarkdown header and chunks derived from the 
# computational components and listdown object to trial-report.rmd.
writeLines(
  paste(c(
    as.character(ld_rmarkdown_header("Simple Trial Report")),
    ld_make_chunks(ld_surv))),
  "trial-report.rmd")

# Render the RMarkdown file to html.
render("trial-report.rmd", quiet = TRUE, pdf_document(keep_tex = TRUE))  
```

# Using \pkg{listdown} in practice for clinical trial reporting

\pagebreak

# Appendix 1: RMarkdown produced in the simple example

```{r eval = FALSE}
---
title: Simple Trial Report
output: html_document
---

\```{r echo=FALSE}
library(gtsummary)
library(survminer)

cc_list <- readRDS("surv-cc.rds")
\```

# Table 1

\```{r echo=FALSE}
cc_list$`Table 1`
\```

# Survival Plots

## Overall

\```{r echo=FALSE}
cc_list$`Survival Plots`$Overall
\```

## Stage

\```{r echo=FALSE}
cc_list$`Survival Plots`$Stage
\```

## Grade

\```{r echo=FALSE}
cc_list$`Survival Plots`$Grade
\```
```

\pagebreak

# Appendix 2: The document produced by the simple example
